---
title: "Data Processing"
author: "Jingjing"
date: '2022-07-13'
output: html_document
---
# Introduction to dplyr

dplyr is an R package in the tidyverse. We can load the package using `install.packages(dplyr)` then get it from the library. 

```{r}
#load required packages and libraries
#install.packages("dplyr")
library(dplyr) #dplyr lets us manipulate data
library(ggplot2)

#preview dataset
head(starwars) #can do comma after object and specify how many rows

```

## Filtering data

We can filter data using `filter()`. We are subsetting our dataset based on the data values. This allows us to subset observations (rows) based on their values/variables (in columns). Subsetting means splitting a full dataset into small subsets. This is useful to limit categories and NA stuff.  

Tips:
* Be sure you spell the column name correctly (and the value name if it's a categorical variable)
* Be sure to use `==` when comparing observations. Remember, `=` is an assignment operator (assign parameters or objects to values)
* We can use `<`, `>`, `==`, `!=`, `<=`, or `>=` to compare numeric or categorical variables (nominal variables are ranked alphabetically while ordinal variables have a built-in rank)


```{r}
filter(starwars, hair_color == "blond" | hair_color == "blonde") 
filter(starwars, hair_color == "blond" & species == "Human" )
filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")

filter(starwars, hair_color %in% c("blond", "blonde", "brown"), species == "Human", height > 75)
filter(starwars, mass != 75 | is.na(mass), name < "Mace") #filters out the rows where the mass is NA

filter(starwars, name < "Mace", mass != 75)
# goes in alphabetical order



filteredData <- filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")


## Find all characters that are shorter than 100 cm
filter(starwars, height < 100, species != "Doid")
View(starwars)



## Which characters were born between the years 100 and 200 (inclusive)
filter(starwars, birth_year %in% c(100:200)) #not good
filter(starwars, birth_year <= 200, birth_year >= 100)
?between()
filter(starwars, between(birth_year, 100, 200)) #inclusive



## Which characters weigh over 100kg but are shorter than 185 cm?
filter(starwars, height < 185 & mass > 100)

## Which characters are missing a hair color?
filter(starwars, is.na(hair_color))



## Missing Values
#NA means the value is unknown, returns unknown
NA > 5 #it doesn't know
NA + 10
NA/2
NA == NA #two unknown values and you don't know if they are equal to each other, doesn't know so returns NA

mass <- NA
mass == 10
is.na(mass) #is.na() figures out whether something is NA, returns TRUE

```

### R Boolean Expressions

**Representing OR** in R is `|`  
<br>
**Representing AND** in R:
* can be `&` if each variable has equal weight
* can use a `,` if want comp to process left to right, and stop processing when reach a FALSE 
<br>
**Representing NOT** in R is just an `!`
**Representing IN** is `%in%`
* put a vector and check is some value matches any one of values in the vector and counts it
* you can use IN when there's one variable and you can compare with a bunch of other ones (use OR is multiple variables)

### Categorical Ordinal Data Filtering
```{r}

##ranked data (categorical ordinal) filter
View(diamonds) #Views dataset
class(diamonds$cut) #gives you the specific type of data
summary(diamonds$cut) #gives you a count of each category or summary statistics (quartiles) if numeric
head(diamonds$cut) #displays first 6 rows, the rankings aare predetermined by creator of dataset

filter(diamonds, cut >= "Very Good")

## Change Ordering categorical data
unique(starwars$eye_color) #tells every unique value there is in a vector
factor(starwars$eye_color, c("red", "orange", "gold", "yellow", "green, yellow", "blue", "black"), ordered = T) #factor() changes things into categories and from nominal to ordinal and the observation unaccounted for turn into <NA>
table(starwars$eye_color) #gives a count of each unique value
```


## Arranging Data
`arrange()` reorders rows. It does not remove any rows.
NA values are always at the end when you order by a column.

```{r}

#lowest to highest birth_year
arrange(starwars, birth_year) 

#highest to lower birth_year
arrange(starwars, desc(birth_year))

#when arranging categorical nominal, goes uppercase --> lowercase, alphabet, if same hair_color, go by default order
arrange(starwars, hair_color)

#multiple columns
arrange(starwars, hair_color, birth_year) #count by hair_color first and if same hair_color --> measure ascending birth_year

### Practice!

## Arrange starwars characters to find the tallest characters and the shortest characters
arrange(starwars, height)
arrange(starwars, desc(height))

## Alphabetize the starwars characters by name
arrange(starwars, name)

## How could you use arrange() to sort all missing values to the start
arrange(starwars, desc(is.na(starwars))) #or list them all out
```

### Selecting Data

We can select certain columns in the dataset. `select()` allows us to retain only certain variables (columns). It doesn't change the order, but it removes columns not named. This is a way to make data anonymous. 

```{r}

select(starwars, hair_color, skin_color, eye_color)

# can select stuff without saying name of columns
select(starwars, hair_color:eye_color) #returns every column between first and last inclusive

select(starwars, -(hair_color:eye_color)) #negative sign gets rid of stuff after it
select(starwars, -hair_color)

starwars_noColor <- select(starwars, -(hair_color:eye_color))
#ggplot(starwars_noColor, aes(x = hair_color)) # error because we removed it

# functions to go along with select()
select(starwars, contains("color")) #pulls every column that contains the word color
select(starwars, ends_with("color"))
select
select(starwars, starts_with("s"), ends_with("color")) # in select() a comma means "OR", in filter(), a comma means "AND"


## Rename Stuff
starwars <- rename(starwars, birth_year = birthYear) 
#starwars <- rename(starwars, birth_year = birthYear) 

## We can use select() to reorder columns
select(starwars, species, species, species) #doesn't repeat columns so only shows columns once
select(starwars, species, everything()) #everything is the rest of them

### Practice!!

## Select out the homeworld and species for the starwars dataset. What question might this subset of variable help answer?
select(starwars, homeworld, species)


## Brainstorm as many ways as possible to select hair_color, eye_color, skin_colors and birth_year



```
### Mutate

`mutate()` adds new columns to the end of the dataset. The new columns created are functions of the previous columns.

Arithmetic operators are all vectorized and will square each row separately (unique for each row) ie: squared, squared-root...

```{r}

starwars_small <- mutate(starwars_small, height_m = height/100)
mutate(starwars_small, bmi = mass/(height_m^2))

#can add multiple columns at once
starwars_small <- select(starwars, height, mass, birth_year)
head(starwars_small)
mutate(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

# to only keep new columns, use transmute
transmute(starwars_small, 
          height_m = height/100, 
          bmi = mass/(height_m^2))


## using aggregate functions (ie: mean, median)
mutate(starwars_small, tot_mass = sum(mass, na.rm = T)) #finds total mass of entire column

mutate(starwars_small, tot_mass = sum(mass, na.rm = T), proportional_mass = mass/tot_mass) #percentage of the total mass each character takes up

```


### Summarize

`summarise()` or `summarize()` is used to summarize and group data. It collapses an entire column of data to a single value. 

```{r}

mutate(starwars, mean_mass = mean(mass, na.rm = T))
summarize(starwars, 
          mean_mass = mean(mass, na.rm = T), 
          mean_height = mean(height, na.rm = T)) #like mutating
#na.rm removes the NA values

#would use summarize() instead of mutate()/transmute() if you want to summarize multiple columns


## Summary of each group
summarise(group_by(starwars, species), 
          mean_mass = mean(mass, na.rm = T)) #get mean_mass of each species

species_masses <- summarize(group_by(starwars, species), 
                          mean_mass = mean(mass, na.rm = T), 
                          count = n()) #n() returns the number of values in each species, number of individuals calculated each time

#the pipe operature %>%
# function(x, y) is the same as x %>% function(y)
#takes everything to the left and follows arrow to become the input of next function
species_masses <- starwars %>%
  group_by(species) %>%
  summarise(mean_mass = mean(mass, na.rm = T), count = n()) %>%
  arrange(desc(mean_mass))

species_masses

#arrange(species_masses, desc(mean_mass))
```

### Sampling a designated number of rows

`sample_n()` allows us to sample a **random** number of rows from our dataset.

```{r}

## 10 random rows
starwars_10rows <- sample_n(starwars, 10) #sample random number of rows, changes each time you run

starwars_10rows #if saved as object, stays the same rows

## 10% of rows, selected randomly
nrow(starwars) #total number of rows is 87
starwars_10percent <- sample_frac(starwars, 0.1)
starwars_10percent #9 rows is 10%

# group_by() with sampling

starwars %>%
  group_by(species)%>%
  sample_n(1) #group_by() finds 1 row per group... species


```



### Slicing a part of the data

We can take a "slice" of our dataset using `slice()` and its related set of functions.This retains the order of our rows and is **NOT** random.

```{r}

slice(starwars, 50:60) #inclusive, slice by index value (row number)

slice_head(starwars, n = 10) #first 10 rows by index
slice_tail(starwars, n = 10) #last 10 rows by index

slice_min(starwars, order_by = mass, n = 10) #rows with lowest 10 heights
slice_max(starwars, order_by = mass, n = 10) #rows with highest 10 masses

```


